### 2013-10-22 msthesis.tex
fixed typos

Recall the matrix $\Psi = \Pi^{1/2}(\Phi - \<\Phi\>)$. The columns of this
matrix, though independent (by choice of independent observables), are not
necessarily orthonormal. However, consider the polar decomposition 
$\Psi = \Q \Z$, where $\Q^t\Q = \I$ and $\Z^2 = \Psi^t\Psi$ is a symmetric
positive semidefinite matrix.\footnote{Recall that the polar decomposition is
  derived from the singular value decomposition, $\Psi = \U\Sigma \V^t$, by
  letting $\Q = \U \V^t$ and $\Z = \V\Sigma \V^t$.}
%% ‘Recall that the polar decomposition is derived from the SVD, \Psi = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
Note that $\Z = (\Psi^t\Psi)^{1/2}$ is nonsingular, so $\Q$ has the form
\[
\Q = \Psi \Z^{-1} = \Psi(\Psi^t\Psi)^{-1/2},
\]
and it is clear that $\ran(\Q) = \ran(\Psi)$. 
Perhaps $\ran(\Q)$ is a useful approximation to the invariant
subspace for $\M$. If $\ran(Q)$ is not itself invariant, we have the error 
matrix $\E =\M\Q - \Q\T$.  Below we show that the
$\T$ which minimizes $\|\E\|_2$ is $\T = \Q^t\M\Q$. 
This yields the following
\begin{theorem}
%Theorem 3.2.2
\label{thm:3.2.2}  
If $\Psi = \Q\Z$ is the polar decomposition of $\Psi$, then the matrix
\[
\T = \Q^t\M\Q = (\Psi^t\Psi)^{-1/2}\Psi^t \M^t \Psi (\Psi^t\Psi)^{-1/2}
\]
minimizes $\|\E\|_2 = \|\M\Q - \Q\T\|_2$.
\end{theorem}
\begin{proof}
We prove the result by establishing the following
\\[6pt]
\underline{Claim:} Given $\M \in \R^{d\times d}$ suppose $Q \in  \R^{d\times d}$
satisfies $\Q^t\Q =\I$. 
Then,
\[
\min_{\T\in \R^{p\times p}} \|\M\Q - \Q\T\|_2 = \|(\I- \Q\Q^t)\M\Q\|_2,
\]
and $\T = \Q^t\M\Q$ is the minimizer.
%
%
% -------------- 18.txt -------------------------------
%
%
The claim is verified by an easy application of the Pythagorean theorem:
For any $\T \in \R^{p\times p}$ we have
\begin{align}
\label{eq:3.1}
\|\M\Q - \Q\T\|^2_2 &= \|\M\Q - \Q\Q^t\M\Q + \Q\Q^t\M\Q - \Q\T\|^2_2 \nonumber\\
&=\|(\I - \Q\Q^t)\M\Q + \Q(\Q^t\M\Q - \T)\|^2_2 \nonumber \\
&=\|(\I - \Q\Q^t)\M\Q\|^2_2 + \|\Q(\Q^t\M\Q - \T)\|^2_2 \\
&=\|(\I - \Q\Q^t)\M\Q\|^2_2 \nonumber
\end{align}
Equality~(\ref{eq:3.1}) holds since $\I - \Q\Q^t$ projects $\M\Q$ onto the
subspace orthogonal to 
$\ran(Q)$. Thus, the two terms in the expression on the right are orthogonal,
and the Pythagorean theorem yields equality. The concluding inequality
establishes that the minimizing $\T$ is that which annihilates the 
second term in~(\ref{eq:3.1}), that is, $\T = \Q^t\M\Q$.

@@

Now, notice that
\begin{equation}
  \label{eq:3.2}
|\rho(\psi_1,\M)| \leq \max |\rho(x,\M)| = \lambda_{\max}(\M)
\end{equation}
% 20.txt
where the max is taken over all nonconstant vectors. Since our interest 
centers on 
$\lambda_{\max}(\M)$, 
we would like a $\psi_1$ that makes the left hand side of~(\ref{eq:3.2}) large. 
This would be achieved if $\psi_1$ were to lie in the space spanned by, say, the
first two eigenvectors of the Markov chain (sometimes referred 
to as the \emph{slowest modes} of the process). However, this subspace spans
only two dimensions of the entire $d$-dimensional space, and it is more likely
that $\psi_1$ only comes close, at best, to lying in the subspace of
interest. Now, given $\psi_1$, a judicious choice for the second dimension $\psi_2$, and
hence $\Psi_2 = [\psi_1, \psi_2]$, would be that which makes 
$\max_{a\neq 0} |\rho(\Psi_2 a)|$ as large as possible. To establish that 
this is indeed the right objective, note the following:
\begin{align}
\label{eq:3.3}
\max_{a\neq 0}|\rho(\Psi_2a,\M)| &= \max_{a\neq 0} \left|\frac{a^t \Psi^t_2 \M \Psi_2 a}{a^t \Psi^t_2 \Psi_2 a}\right|\nonumber\\
&= \max_{x\in \ran(\Psi_2)}|\rho(x,\M)|\nonumber\\
&= \max|\rho(x,\M)|\\
&= \lambdamax\nonumber
\end{align}
Again, the max on the right side of~(\ref{eq:3.3}) is over nonconstant vectors. 
In other words, we wish to chose
$\Psi_2 = [\psi_1, \psi_2]$ so that there is a vector $a \in \R^2$ making 
$|\rho(\Psi_2 a,\M)|$ close to $\lambdamax$.

@@

\section{The General Lanczos Algorithm}
% Sec 4.1
\label{sec:gener-lancz-algor}
Let $\psip{k} =  \M^k \psi$ and consider the $d\times d$ matrix
\[
\Psi = [\psi, \M\psi, \dots, \M^{d-1}\psi] = [\psip{0}, \psip{1}, \dots, \psip{d-1}]
\]
Notice that $\M\Psi = [\psip{1}, \psip{2}, \dots, \psip{d-1}, \M^d \psi]$,
and assuming for the moment that $\Psi$ is nonsingular,
we can compute the vector $h = -\Psi^{-1}\M^d \psi$. Thus,
\begin{equation}
\label{eq:10000}
\M\Psi = \Psi [e_2,\dots,e_d,-h],  
\end{equation}
where $e_j$ is the column vector with 1 in
the $j$th position and zeros elsewhere. We define $H = [e_2,\dots,e_d,-h]$,
so~(\ref{eq:10000}) becomes $\M\Psi = \Psi H$.
%
%
% ----------------- 23.txt ----------------------------------------------
%
%
Equivalently,
\begin{equation}
  \label{eq:4.1}
H = 
\begin{pmatrix}
0 & 0      & \cdots & 0 & -h_1 \\
1 & 0      & \cdots & 0 & -h_2\\
  & \ddots &        &   & \vdots \\
  &        & \ddots & 0 & -h_{d-1}\\
  &        &        & 1 & -h_{d}
\end{pmatrix} = \Psi^{-1} \M \Psi.
\end{equation}
$\H$ is a \emph{companion matrix} which means that its characteristic polynomial
is $p(x) = x^d + \sum_{i=1}^d h_i x^{i-1}$.
Since $\H$ is similar to $\M$, finding the eigenvalues of $\M$ is equivalent to
finding the roots of $p(x)$. However, this is of little practical use since
finding $h$, constructing $p(x)$, and finding its roots is 
probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
derives from its \emph{upper Hessenberg form}. We exploit this property below.

Let $\Psi = \Q\mR$ be the QR decomposition of $\Psi$. Since $\Psi$ is assumed nonsingular,
\[
\Psi_d^{-1}\M \Psi_d = (\mR^{-1}\Q^t)\M(\Q\mR) = \H
\]
Therefore, $\Q^t\M\Q = \mR\H\mR^{-1}$.  Let $\T \mR\H\mR^{-1}$.  Then, since $\mR$
and $\mR^{-1}$ are both upper triangular and $\H$ is upper Hessenberg $\T = \mR\H\mR^{-1}$ is also upper
Hessenberg. Furthermore, since $\M$ is symmetric, it is clear that 
$\T^t =\Q^t\M\Q = T$. Thus, $\T$ is both upper Hessenberg and symmetric. Therefore,
$\T$ is tridiagonal and we can write it as follows:
\begin{equation}
  \label{eq:4.2}
\T = 
\begin{pmatrix}
\alpha_1 & \beta_1 &      0 & \cdots & 0 \\
\beta_1  & \alpha_2 & \ddots &        & \vdots\\
         & \ddots & \ddots & \ddots   &\\
\vdots   &       &  \ddots & \ddots & \beta_{d-1}\\
0  &   \cdots    &   & \beta_{d-1} &\alpha_d
\end{pmatrix}.
\end{equation}
%
%
% ------------ 24.txt ----------------------------------------------
%
%
Equating columns $j$ on both sides of $\M\Q = \Q\T$ yields
\begin{equation}
\label{eq:4.3}
\M q_j = \beta_{j-1}q_{j-1}+\alpha_j q_j + \beta_j q_{j+1}
\end{equation}
Since the columns of $\Q$ are orthonormal, multiplying both sides 
of~(\ref{eq:4.3})
by $q_j$ and $q_{j+1}$ yields
$\alpha_j = q_j^t \M q_j$ and $\beta_j = q_{j+1}^t \M q_j$.
The foregoing justifies what is called the \emph{Lanczos algorithm}, which is performed as follows:
         \begin{algorithm}
           \caption{The Lanczos algorithm for partial reduction to symmetric
             tridiagonal form.}
           \begin{algorithmic}
             \STATE $q_1 = \psi/\|\psi\|_2$, $\beta_0 = 0$, $q_0= 0$
             \FOR{$j = 1, \dots, p$}
             \STATE $z \leftarrow \M q_j$
             \STATE $\alpha_j \leftarrow q_j^tz$
             \STATE $z \leftarrow z - \alpha_j q_j - \beta_{j-1} q_{j-1}$
             \STATE $\beta_j \leftarrow \|z\|_2$
             \IF{$\beta_j=0$}
             \STATE return
             \ENDIF
             \STATE $q_{j+1} = z/\beta_j$
             \ENDFOR
           \end{algorithmic}
         \end{algorithm}
The $q_j$ computed by the Lanczos algorithm are often called the \emph{Lanczos vectors}. If the
loop in the algorithm is terminated because $\beta_{k}= 0$, this indicates that
an exact invariant subspace has been computed, and is given by 
$\ran\{q_1, \dots, q_k\}$. Otherwise, we usually halt the algorithm after 
$p$ steps, in which case the algorithm converges to approximations of at most
  $p$ eigenvalues.\footnote{A Detailed discussion of such convergence (and
    misconvergence) is given in~\cite{Demmel:1997}.}

@@

The expressions above involve the higher order Lanczos vectors $q_j$. In the present context,
these vectors will not be readily available. However, another recursion---this time relating forms
involving $q_{j+1}$ to forms involving $q_j$ and $q_{j-1} $---will make these
expressions accessible by induction on our initial observable. The general forms
of interest are $q_{j+1}^t\M^k q_{j+1}$ and $q^t_{j+1}\M^k q_j$. Again, we 
begin with the simpler of the two:
\begin{align*}
  q_{j+1}^tM^kq_j &= 
 \frac{1}{\beta_j}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k q_j\\
 \frac{1}{\beta_j}(q_j^t\M^{k+1} q_j - \alpha_j q_j^t\M^{k} q_j -\beta_{j-1} q_j^t\M^k).
\end{align*}
The second is,
\begin{align}
\label{eq:4.12}
  q_{j+1}^tM^kq_{j+1} 
&=  \frac{1}{\beta_j^2}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k 
  (\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})\nonumber\\
&=  \frac{1}{\beta_j^2}(q_j^t\M^{k+2} q_j + \alpha^2_jq_j^t\M^{k} q_j 
  + \beta^2_{j-1} q_{j-1}^t\M^k q_{j-1} - 2\alpha_jq_j^t\M^{k+1} q_j \nonumber\\
& \qquad    
    + 2\alpha_j \beta_{j-1}q_j^t\M^k q_{j-1}
    - 2\beta_{j-1}q_j^t \M^{k+1}q_{j-1}).
\end{align}
The point of all this is that each pair of coefficients $(\alpha_{j+1}, \beta_j)$
can be expressed as a function of the covariance between 
$\phi(n)$ and $\phi(n+ m)$, for $m = 0, \dots, 2j+1$. Since the expressions
are based on four recurrence relations, they become quite complicated as $j$ gets large. However,
defining these recursions in a computer program is trivial, as we demonstrate
in Section~\ref{sec:computations} % Sec 5.4
by applying the foregoing to a particular problem.
%
%
% ----------------- 30.txt ------------------------------------------------------
%
%

>>>>

Recall the matrix $\Psi = \Pi^{1/2}(\Phi - \<\Phi\>)$. The columns of this
matrix, though independent (by choice of independent observables), are not
necessarily orthonormal. However, consider the polar decomposition 
$\Psi = \Q \Z$, where $\Q^t\Q = \I$ and $\Z^2 = \Psi^t\Psi$ is a symmetric
positive semidefinite matrix.\footnote{Recall that the polar decomposition is
  derived from the singular value decomposition, $\Psi = \U\Sigma \V^t$, by
  letting $\Q = \U \V^t$ and $\Z = \V\Sigma \V^t$.}
%% ‘Recall that the polar decomposition is derived from the SVD, \Psi = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
Note that $\Z = (\Psi^t\Psi)^{1/2}$ is nonsingular, so $\Q$ has the form
\[
\Q = \Psi \Z^{-1} = \Psi(\Psi^t\Psi)^{-1/2},
\]
and it is clear that $\ran(\Q) = \ran(\Psi)$. 
Perhaps $\ran(\Q)$ is a useful approximation to an invariant
subspace for $\M$. If $\ran(\Q)$ is not itself invariant, we have the error 
matrix $\E =\M\Q - \Q\T$.  Below we show that the
$\T$ which minimizes $\|\E\|_2$ is $\T = \Q^t\M\Q$. 
This yields the following
\begin{theorem}
%Theorem 3.2.2
\label{thm:3.2.2}  
If $\Psi = \Q\Z$ is the polar decomposition of $\Psi$, then the matrix
\[
\T = \Q^t\M\Q = (\Psi^t\Psi)^{-1/2}\Psi^t \M^t \Psi (\Psi^t\Psi)^{-1/2}
\]
minimizes $\|\E\|_2 = \|\M\Q - \Q\T\|_2$.
\end{theorem}
\begin{proof}
We prove the result by establishing the following
\\[6pt]
\underline{Claim:} Given $\M \in \R^{d\times d}$ suppose $Q \in  \R^{d\times d}$
satisfies $\Q^t\Q =\I$. 
Then,
\[
\min_{\T\in \R^{p\times p}} \|\M\Q - \Q\T\|_2 = \|(\I- \Q\Q^t)\M\Q\|_2,
\]
and $\T = \Q^t\M\Q$ is the minimizer.
%
%
% -------------- 18.txt -------------------------------
%
%
The claim is verified by an easy application of the Pythagorean Theorem:
For any $\T \in \R^{p\times p}$ we have
\begin{align}
\label{eq:3.1}
\|\M\Q - \Q\T\|^2_2 &= \|\M\Q - \Q\Q^t\M\Q + \Q\Q^t\M\Q - \Q\T\|^2_2 \nonumber\\
&=\|(\I - \Q\Q^t)\M\Q + \Q(\Q^t\M\Q - \T)\|^2_2 \nonumber \\
&=\|(\I - \Q\Q^t)\M\Q\|^2_2 + \|\Q(\Q^t\M\Q - \T)\|^2_2 \\
&=\|(\I - \Q\Q^t)\M\Q\|^2_2 \nonumber
\end{align}
Equality~(\ref{eq:3.1}) holds since $\I - \Q\Q^t$ projects $\M\Q$ onto the
subspace orthogonal to 
$\ran(\Q)$. Thus, the two terms in the expression on the right are orthogonal,
and the Pythagorean theorem yields equality. The concluding inequality
establishes that the minimizing $\T$ is that which annihilates the 
second term in~(\ref{eq:3.1}), that is, $\T = \Q^t\M\Q$.

@@

Now, notice that
\begin{equation}
  \label{eq:3.2}
|\rho(\psi_1)| \leq \max |\rho(x)| = \lambda_{\max}(\M)
\end{equation}
% 20.txt
where the max is taken over all nonconstant vectors. Since our interest 
centers on 
$\lambda_{\max}(\M)$, 
we would like a $\psi_1$ that makes the left hand side of~(\ref{eq:3.2}) large. 
This would be achieved if $\psi_1$ were to lie in the space spanned by, say, the
first two eigenvectors of the Markov chain (sometimes referred 
to as the \emph{slowest modes} of the process). However, this subspace spans
only two dimensions of the entire $d$-dimensional space, and it is more likely
that $\psi_1$ only comes close, at best, to lying in the subspace of
interest. Now, given $\psi_1$, a judicious choice for the second dimension $\psi_2$, and
hence $\Psi_2 = [\psi_1, \psi_2]$, would be that which makes 
$\max_{a\neq 0} |\rho(\Psi_2 a)|$ as large as possible. To establish that 
this is indeed the right objective, note the following:
\begin{align}
\label{eq:3.3}
\max_{a\neq 0}|\rho(\Psi_2a)| &= \max_{a\neq 0} \left|\frac{a^t \Psi^t_2 \M \Psi_2 a}{a^t \Psi^t_2 \Psi_2 a}\right|\nonumber\\
&= \max_{x\in \ran(\Psi_2)}|\rho(x)|\nonumber\\
&= \max|\rho(x)|\\
&= \lambdamax\nonumber
\end{align}
Again, the max on the right side of~(\ref{eq:3.3}) is over nonconstant vectors. 
In other words, we wish to chose
$\Psi_2 = [\psi_1, \psi_2]$ so that there is a vector $a \in \R^2$ making 
$|\rho(\Psi_2 a)|$ close to $\lambdamax$.

@@

\section{The General Lanczos Algorithm}
% Sec 4.1
\label{sec:gener-lancz-algor}
Let $\psip{k} =  \M^k \psi$ and consider the $d\times d$ matrix
\[
\Psi = [\psi, \M\psi, \dots, \M^{d-1}\psi] = [\psip{0}, \psip{1}, \dots, \psip{d-1}]
\]
Notice that $\M\Psi = [\psip{1}, \psip{2}, \dots, \psip{d-1}, \M^d \psi]$,
and assuming for the moment that $\Psi$ is nonsingular,
we can compute the vector $h = -\Psi^{-1}\M^d \psi$. Thus,
\begin{equation}
\label{eq:10000}
\M\Psi = \Psi [e_2,\dots,e_d,-h],  
\end{equation}
where $e_j$ is the column vector with 1 in
the $j$th position and zeros elsewhere. We define 
$\H = [e_2,\dots,e_d,-h]$,
so~(\ref{eq:10000}) becomes $\M\Psi = \Psi \H$.
%
%
% ----------------- 23.txt ----------------------------------------------
%
%
Equivalently,
\begin{equation}
  \label{eq:4.1}
\H = 
\begin{pmatrix}
0 & 0      & \cdots & 0 & -h_1 \\
1 & 0      & \cdots & 0 & -h_2\\
  & \ddots &        &   & \vdots \\
  &        & \ddots & 0 & -h_{d-1}\\
  &        &        & 1 & -h_{d}
\end{pmatrix} = \Psi^{-1} \M \Psi.
\end{equation}
$\H$ is a \emph{companion matrix} which means that its characteristic polynomial
is $p(x) = x^d + \sum_{i=1}^d h_i x^{i-1}$.
Since $\H$ is similar to $\M$, finding the eigenvalues of $\M$ is equivalent to
finding the roots of $p(x)$. However, this is of little practical use since
finding $h$, constructing $p(x)$, and finding its roots is 
probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
derives from its \emph{upper Hessenberg form}. We exploit this property below.

Let $\Psi = \Q\mR$ be the QR decomposition of $\Psi$. Since $\Psi$ is assumed nonsingular,
\[
\Psi_d^{-1}\M \Psi_d = (\mR^{-1}\Q^t)\M(\Q\mR) = \H
\]
Therefore, $\Q^t\M\Q = \mR\H\mR^{-1}$.  Let $\T \mR\H\mR^{-1}$.  Then, since $\mR$
and $\mR^{-1}$ are both upper triangular and $\H$ is upper Hessenberg $\T = \mR\H\mR^{-1}$ is also upper
Hessenberg. Furthermore, since $\M$ is symmetric, it is clear that 
$\T^t =\Q^t\M\Q = T$. Thus, $\T$ is both upper Hessenberg and symmetric. Therefore,
$\T$ is tridiagonal and we can write it as follows:
\begin{equation}
  \label{eq:4.2}
\T = 
\begin{pmatrix}
\alpha_1 & \beta_1 &      0 & \cdots & 0 \\
\beta_1  & \alpha_2 & \ddots &        & \vdots\\
         & \ddots & \ddots & \ddots   &\\
\vdots   &       &  \ddots & \ddots & \beta_{d-1}\\
0  &   \cdots    &   & \beta_{d-1} &\alpha_d
\end{pmatrix}.
\end{equation}
%
%
% ------------ 24.txt ----------------------------------------------
%
%
Equating columns $j$ on both sides of $\M\Q = \Q\T$ yields
\begin{equation}
\label{eq:4.3}
\M q_j = \beta_{j-1}q_{j-1}+\alpha_j q_j + \beta_j q_{j+1}
\end{equation}
Since the columns of $\Q$ are orthonormal, multiplying both sides 
of~(\ref{eq:4.3})
by $q_j$ and $q_{j+1}$ yields
$\alpha_j = q_j^t \M q_j$ and $\beta_j = q_{j+1}^t \M q_j$.
The foregoing justifies what is called the \emph{Lanczos algorithm}, which is performed as follows:
         \begin{algorithm}
           \caption{The Lanczos algorithm for partial reduction to symmetric
             tridiagonal form.}
           \begin{algorithmic}
             \STATE $q_1 = \psi/\|\psi\|_2$, $\beta_0 = 0$, $q_0= 0$
             \FOR{$j = 1, \dots, p$}
             \STATE $z \leftarrow \M q_j$
             \STATE $\alpha_j \leftarrow q_j^tz$
             \STATE $z \leftarrow z - \alpha_j q_j - \beta_{j-1} q_{j-1}$
             \STATE $\beta_j \leftarrow \|z\|_2$
             \IF{$\beta_j=0$}
             \STATE return
             \ENDIF
             \STATE $q_{j+1} = z/\beta_j$
             \ENDFOR
           \end{algorithmic}
         \end{algorithm}

The $q_j$ computed by the Lanczos algorithm are often called the \emph{Lanczos vectors}. If the
loop in the algorithm is terminated because $\beta_{k}= 0$, this indicates that
an exact invariant subspace has been computed, and is given by 
$\ran\{q_1, \dots, q_k\}$. Otherwise, we usually halt the algorithm after 
$p$ steps, in which case the algorithm converges to approximations of at most
  $p$ eigenvalues.\footnote{A Detailed discussion of such convergence (and
    misconvergence) is given in~\cite{Demmel:1997}.}

@@

The expressions above involve the higher order Lanczos vectors $q_j$. In the present context,
these vectors will not be readily available. However, another recursion---this time relating forms
involving $q_{j+1}$ to forms involving $q_j$ and $q_{j-1} $---will make these
expressions accessible by induction on our initial observable. The general forms
of interest are $q_{j+1}^t\M^k q_{j+1}$ and $q^t_{j+1}\M^k q_j$. Again, we 
begin with the simpler of the two:
\begin{align*}
  q_{j+1}^t\M^kq_j &= 
 \frac{1}{\beta_j}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k q_j\\
&=  \frac{1}{\beta_j}(q_j^t\M^{k+1} q_j - \alpha_j q_j^t\M^{k} q_j -\beta_{j-1} q_j^t\M^k).
\end{align*}
The second is,
\begin{align}
\label{eq:4.12}
  q_{j+1}^t\M^kq_{j+1} 
&=  \frac{1}{\beta_j^2}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k 
  (\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})\nonumber\\
&=  \frac{1}{\beta_j^2}(q_j^t\M^{k+2} q_j + \alpha^2_jq_j^t\M^{k} q_j 
  + \beta^2_{j-1} q_{j-1}^t\M^k q_{j-1} - 2\alpha_jq_j^t\M^{k+1} q_j \nonumber\\
& \qquad    
    + 2\alpha_j \beta_{j-1}q_j^t\M^k q_{j-1}
    - 2\beta_{j-1}q_j^t \M^{k+1}q_{j-1}).
\end{align}
The point of all this is that each pair of coefficients $(\alpha_{j+1}, \beta_j)$
can be expressed as a function of the covariance between 
$\phi(n)$ and $\phi(n+ m)$, for $m = 0, \dots, 2j+1$. Since the expressions
are based on four recurrence relations, they become quite complicated as $j$ gets large. However,
defining these recursions in a computer program is trivial, as we demonstrate
in Section~\ref{sec:computations} % Sec 5.4
by applying the foregoing to a particular problem.
%
%
% ----------------- 30.txt ------------------------------------------------------
%
%
